Master's Project Journal

### 2019-04-24 thesis/Background.tex
DONEgit add -u!

\subsection{Digital Loop Filter}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../simple_pi}
	\caption[Basic \ac{PI} controller architecture]{Basic \ac{PI} controller architecture.}
	\label{fig:my_simple_pi}
\end{figure}
The \acl{LF} in \iac{ADPLL} can be implemented as \iac{PI} controller, as only a low-pass filter is required, such as that in Figure \ref{fig:my_simple_pi}. In the case of a node in \iac{ADPLL} network the input of this filter is a weighted average of the phase difference relative to the neighbouring local clocking areas. In a digital system, the proportional section can be implemented by a simple multiplier, whereas the integral path is constructed by the delayed addition of error multiplied by the integral gain to the sum of past values. This delayed accumulation can be easily implemented by an accumulator to which the current value of the multiplication is added each cycle, and as a result the system has an infinite impulse response. The value of these gains determine the response and stability of the \ac{ADPLL} network. The transfer function of such a controller is given by \cite{shan2014phd}:
\begin{equation*}
	H(z) = \alpha + \beta\frac{1}{1-z^{-1}} = \frac{(\alpha + \beta) - \alpha z^{-1}}{1-z^{-1}}\vspace{-0.2cm}
\end{equation*}
It has been found by Koskin \textit{et al} (2018), that stable operation can be achieved when the integral gain, $k_i$, is less than half the proportional gain, $k_p$ \cite{koskin2018generation}. In the same study a range of values was found which would produce low jitter operation of the network. As these values are all less than one, the filter must implement fixed point arithmetic in an effort to maintain the simplicity of the \ac{LF}, rather than incurring the penalty of floating point calculations.

@@

\section{\acs{ADPLL} Performance Characterisation}
As already stated, the goal of a clock distribution system is to synchronise clocking events in all areas of the chip. The effectiveness of this synchronisation is characterised by two main metrics, jitter and skew, which together describe the distribution of clocking events. Skew is the average time delay between a clocking event in one area of the chip and that of a reference event. It can be easily measured by computing the average value of this delay. Jitter has a number of definitions depending on how it is measured, but at its most basic, it is the standard deviation of the time delays with respect to the reference clocking edge.

The simplest form of clock performance characterisation is on \iac{C2C} basis, in which the reference is the previous edge of the signal under test itself. Here the standard deviation of the individual delays gives the jitter of the signal. As the previous edge of the generated signal is used as the reference, and thereby the delay is the period of the signal, the mean value is no longer the skew, but instead can be used to compute the centre frequency over the time interval. For a more informative measurement the \ac{TIE} can be computed, which takes into account another signal as the reference. \ac{TIE} is calculated by comparing the delay between the signal in question and a reference that is treated as ideal. The mean value gives the relative skew between the two signals and, once more, the standard deviation of the measurements gives the jitter with respect to this reference.

Both above forms of measurement assume a large number of sequential measurements, however, there are other ways that jitter and skew can be calculated \cite{AN10007}. Phase jitter is an important characteristic for communications systems as it can be used to calculate phase noise, important to ensure spurious emissions are within legal requirements. Long term, or accumulated, jitter represents the cumulative effect of jitter on the signal over several cycles. Long term jitter in particular affects \acs*{RADAR} as it will manifest itself as a Doppler shift in the return signal. For \iac{PLL} network, with an appropriately designed filter, the long term jitter should be zero so long as the reference signal is stable. %TODO ask Brian about this one

>>>>

\subsection{Digital Loop Filter}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../simple_pi}
	\caption[Basic \acs{PI} controller architecture]{Basic \acs{PI} controller architecture.}
	\label{fig:my_simple_pi}
\end{figure}
The \acl{LF} in \iac{ADPLL} can be implemented as \iacf{PI} controller, as only a low-pass filter is required, such as that in Figure \ref{fig:my_simple_pi}. In the case of a node in \iac{ADPLL} network the input of this filter is a weighted average of the phase difference relative to the neighbouring local clocking areas. In a digital system, the proportional section can be implemented by a simple multiplier, whereas the integral path is constructed by the delayed addition of error multiplied by the integral gain to the sum of past values. This delayed accumulation can be easily implemented by an accumulator to which the current value of the multiplication is added each cycle, and as a result the system has an infinite impulse response. The value of these gains determine the response and stability of the \ac{ADPLL} network. The transfer function of such a controller is given by \cite{shan2014phd}:
\begin{equation*}
	H(z) = \alpha + \beta\frac{1}{1-z^{-1}} = \frac{(\alpha + \beta) - \alpha z^{-1}}{1-z^{-1}}\vspace{-0.2cm}
\end{equation*}
It has been found by Koskin \textit{et al} (2018), that stable operation can be achieved when the integral gain, $k_i$, is less than half the proportional gain, $k_p$ \cite{koskin2018generation}. In the same study a range of values was found which would produce low jitter operation of the network. As these values are all less than one, the filter must implement fixed point arithmetic in an effort to maintain the simplicity of the \ac{LF}, rather than incurring the penalty of floating point calculations.

@@

\section{\acs{ADPLL} Performance Characterisation}
The goal of a clock distribution system is to synchronise clocking events in all areas of the chip. The effectiveness of this synchronisation is characterised by two main metrics, jitter and skew, which together describe the distribution of clocking events. Skew is the average time delay between a clocking event in one area of the chip and that of a reference event. It can be easily measured by computing the average value of this delay. Jitter has a number of definitions depending on how it is measured, but at its most basic, it is the standard deviation of the time delays with respect to the reference clocking edge.

The simplest form of clock performance characterisation is on \iac{C2C} basis, in which the reference is the previous edge of the signal under test itself. Here the standard deviation of the individual delays gives the jitter of the signal. As the previous edge of the generated signal is used as the reference, and thereby the delay is the period of the signal, the mean value is no longer the skew, but instead can be used to compute the centre frequency over the time interval. For a more informative measurement, the \ac{TIE} can be computed, which takes into account another signal as the reference. \ac{TIE} compares the delay between the signal in question and a reference that is treated as ideal. The mean value gives the relative skew between the two signals and, once more, the standard deviation of the measurements gives the jitter with respect to this reference.

There are a number of other ways that jitter and skew can be calculated \cite{AN10007}. Phase jitter is important for communications systems, for the calcuation of phase noise, used to ensure spurious emissions are within legal requirements. Long term, or accumulated, jitter represents the cumulative effect of jitter on the signal over several cycles. Long term jitter in particular affects \acs*{RADAR} as it will manifest itself as a Doppler shift in the return signal. For \iac{PLL} network, with an appropriately designed filter, the long term jitter should be zero so long as the reference signal is stable.

### 2019-04-24 thesis/Conclusion.tex
DONEgit add -u!

\section{Conclusions}
\section{Suggested Future Work}
\subsection{\acl{TDL} Characterisation}
\acreset{TDL}
As previously mentioned, characterisation of the \ac{TDL} was not carried out due to the time it would have taken to perform the required measurements and tests. This characteristion can be accomplished through the use of an off-chip tunable delay. Off-chip is unfortunately a requirement, as the only delays available of the order required to test the step sizes are the variable delay of the inverters themselves. A resistor \& capacitor delay circuit can be used to perform these test, and the output of the \ac{PFD} analysed on a mixed-signal oscilloscope alongside the delayed and reference waveforms. The use of an off-chip delay element necessitates careful construction/cailbration of the measurement environment to eliminate the delay stemming from the signal path through input/output buffers as it leaves the chip.

\subsection{Increased Network Size}
A 4x4 network was a stretch-goal of this project, depending on the time remaining after once the minumim goal of a 3x3 network had been reached. Unfortuantely due to problems with the clocked phase detector which resulted in mode-locking, incorrect \ac{RO} inverter selection logic and the projected time requirement to ensure alignment of each \ac{RO} this was not achieved. A 4x4 network would better explain some dynamics, such as the skew growth as the \ac{ADPLL} indices increase. However. progressing beyond a 3x3 network means that less than $20\%$ of \ac{ADPLL} output signals can be examined at any given point in time. As such, it may be useful to develop a more advanced measurement set-up that can either capture more signals simultaneously, or implement a measurement system that makes more efficient use of the available probes.

\subsection{\acs{FPGA} Clocked, Linear Period \acs{DCO}}
\acreset{FPGA,DCO}
While this design of \ac{DCO} was proposed earlier in this thesis, it was not implemented in any \ac{ADPLL}s, as it was only realised that the output waveform did not matter, just the location of the rising edges was not made until later in the project, once the pitfall encountered in the clocked \ac{PFD} design was resolved. Until this point the non-square wave output had eliminated the design from consideration. However, as in an \ac{ASIC} linear period designs are easier to implement, it would be ideal to have the benefit of frequency scaled hardware simulations with a \ac{DCO} that more closely relates the the eventual system.

\subsection{Procedural Network Instantiation}
This suggestion is directed at making the user experience simpler. As implemented, networks based on \ac{ADPLL}s containing \ac{RO}s are implemented in separate files to their \ac{FPGA} clocked \ac{DCO} based counterparts as are the networks of difference sizes. At a network size of 2x2 this was not a problem, but at the scale of 3x3 network the modules became somewhat cumbersome to work with, with a large number of interconnects. For 3x3 and larger networks it would be expedient to create the network with a \texttt{generate} statement, as this would reduce the number of lines of code and the amount of named interconnects required. Instead the user could set a number of configuration parameters and the network would be created accordingly. This could even allow for the use of different \ac{ADPLL} types in the same network with ease. Unfortunately the tasks of floorplanning and oscillator alignment would still be left to the user to perform.

%TODO cahracterise tdl

>>>>

\section{Conclusions}
\acuse{FPGA,ADPLL,RO,TDL,DCO,ASIC,PFD,EDA}
The objective of this project was to create \iacs{FPGA} based platform that could be used to prototype and model \acs{ADPLL} networks in a cost and time effective manner. The main effort of the project was to analyse the ways in which the components of \iacs{ADPLL} network could be designed, and then selecting a number of the most appropriate designs from this pool. A number of \acs{ADPLL} designs were then created using the selected components in different combinations for the purposes of comparison. Once these \acs{ADPLL} designs had been validated, they were implemented in networks of a 2x2 and 3x3 Cartesian grid topology. Measurements and performance benchmarks were then carried out on these \acs{ADPLL} networks, and the results used to verify claims made regarding the intended use cases of the individual \acs{ADPLL} designs. Finally, various small modifications to the \acs{ADPLL} designs were made, and their impact on system performance identified.

The project resulted in a platform comprising three \acs{ADPLL} designs suited for different use cases. As evidenced by the investigation into the impact of minor variations, this system fulfils the objective that it be suitable for the testing of architectural changes. Realistic, though exaggerated, variation due to implementation makes the inverter based designs ideal for the testing of new control schemes or performance enhancing modules. Simultaneously, this makes designs based on \acsp{RO} ideal for the confirmation of either theoretical or simulated results, in a platform that suffers from realistic implementation based variation, and non-ideal effects, such as power supply noise, that are difficult to simulate well. The presence of \acs{FPGA} clocked \acsp{DCO} and \acsp{PFD} makes the emulation of mixed-signal circuits, at a scaled down frequency, for the purposes of design verification possible.

As the platform is designed with extensibility in mind, rapid changes in the design are possible, with the lengthy and resource consuming tape-out process that would otherwise be required for a test in hardware, cut down to the time taken for the \ac{EDA} tool of choice to generate a configuration file. Whereas tape-out may be a costly process, for a figure in the $100$s of Euro, an \acs{FPGA} can be acquired that will bridge the gap between software simulations and implementation in custom silicon, by providing a hardware-based simulation, modelling and validation platform.

\section{Suggested Future Work}
\subsection{\acl{TDL} Characterisation}
As previously mentioned, characterisation of the \ac{TDL} was not carried out due to the time it would have taken to perform the required measurements and tests. This characteristion can be accomplished through the use of an off-chip tunable delay. Off-chip is unfortunately a requirement, as the only delays available of the order required to test the step sizes are the variable delay of the inverters themselves. A resistor \& capacitor delay circuit can be used to perform these tests, and the output of the \ac{PFD} analysed on a mixed-signal oscilloscope, alongside the delayed and reference waveforms. The use of an off-chip delay element necessitates careful construction/cailbration of the measurement environment to eliminate the delay stemming from the signal path through input/output buffers as it leaves the chip.

\subsection{Increased Network Size}
If timed had permitted, 4x4 network was to be implemented as part of this project. Unfortuantely due to problems with the clocked phase detector  (which resulted in mode-locking), incorrect \ac{RO} inverter selection logic and the projected time requirement to ensure alignment of 16 \acp{RO} this was not achieved. A 4x4 network would better explain some dynamics, such as the skew growth as the \ac{ADPLL} indices increase. However, progressing beyond a 3x3 network means that less than $20\%$ of \ac{ADPLL} output signals can be examined at any one point in time. Therefore, it may be useful to develop a more advanced measurement set-up, that can either capture more signals simultaneously, or implement a measurement system that makes more efficient use of the available probes.

\subsection{\acs{FPGA} Clocked, Linear Period \acs{DCO}}
While this type of \ac{DCO} was proposed earlier in this thesis, it was not implemented in any \acp{ADPLL}. At the time the \acp{ADPLL} were designed, a square wave output was required, something this design cannot accomplish. It was not until later in the project, once the pitfall encountered in the clocked \ac{PFD} design was resolved, that a non square wave signal would have been possible. Until this point, the output waveform had eliminated the design from consideration. However, as in \iac{ASIC}, linear period designs are easier to implement. It would be ideal to have the benefit of frequency scaled hardware simulations with \iac{DCO} that more closely relates to the eventual system.

\subsection{Procedural Network Instantiation}
This suggestion is directed at making the user experience simpler. As implemented, networks based on \ac{ADPLL}s containing \ac{RO}s are implemented in separate files to their \ac{FPGA} clocked \ac{DCO} based counterparts, as are the networks of difference sizes. At a network size of 2x2, this was not a problem, but with a 3x3 network, the modules became somewhat cumbersome to work with. This is due to the large number of interconnects. For 3x3 and larger networks, it would be expedient to create the network with a \texttt{generate} statement, as this would reduce the number of lines of code and the number of individually named interconnects required. Instead the user could set a number of configuration parameters and the network would be created accordingly. This could even allow for the use of different \ac{ADPLL} types in the same network with ease. Unfortunately, the tasks of floorplanning and oscillator alignment would still be left to the user to perform.

### 2019-04-24 thesis/FPGABasedADPLLs.tex
DONEgit add -u!

\section{Chapter Overview}
The first step in creating \iac{FPGA} based network of \acp{ADPLL} is the design of the \ac{ADPLL} itself, which will be addressed in this chapter. The nature of \iac{FPGA} necessitates compromises in the design of a certain blocks, limiting direct transfer from \ac{ASIC} designs. In this chapter the potential designs for each individual block, or module, investigated will be explained and the case for their selection in \iac{FPGA} based \ac{ADPLL} made. A number of blocks have implemented by purely digital circuitry and as such can be transferred in their entirety from \iac{FPGA} and vice versa. However, those that will be used to emulate mixed-signal circuitry, such as the \ac{DCO} and \ac{PFD}, will be examined in greater detail. \Iac{RTL} diagram of the \ac{ADPLL} topology that will be used in the following sections is shown in Figure \ref{fig:network_adpll}.
\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{../network_adpll}
	\caption{\ac{ADPLL} designed for use in a Cartesian grid network.}
	\label{fig:network_adpll}
\end{figure}

@@

\subsection{\acs{FIR} \acl{LF}}
The \acl{FIR} implementation of the \acl{LF} will take the form presented in Figure \ref{fig:fir_pi} (a), and consists series of time delays, or ``taps'', the output of which is multiplied by a coefficient. The result of each multiplication is added together to form the filter output. Various tools, Matlab for example, can be used to compute the coefficients and number of taps required to obtain a specific filter response. Using \iac{FIR} filter is advantageous as it is always stable due to the lack of a feedback path, however a large number of taps are required to obtain a frequency response akin to that of the \ac{IIR} implementation, thus requiring an equal number of multipliers and $taps-1$ adders. Compared to \iac{IIR}, which requires just two adders, two multipliers and a flip-flop this, has a much greater hardware footprint. It is possible to condense the \ac{FIR} multiplication and addition to a single multiplier and adder, depicted in \ref{fig:fir_pi} (b), however this will take as many clock cycles to complete as there are taps. This is not viable in an \ac{ADPLL} as the filter must produce its output in a single clock cycle. In the case of any coefficient symmetry, addition can be performed prior to multiplication, thus reducing the number of multipliers for each identical coefficient.
\begin{figure}[h]
	\centering
	\subfloat[Typical FIR Filter]{{\includegraphics[width=0.6\textwidth]{../FIR_1} }}\\
	\subfloat[Single Multiplier FIR Filter]{{\includegraphics[width=0.6\textwidth]{../FIR_2} }}%
	\caption[\ac{FIR} \ac{PI} loop filter RTL diagrams]{\ac{FIR} \ac{PI} loop filter RTL diagrams.}
	\label{fig:fir_pi}
\end{figure}

@@

As the scaling factor used in this multiplier is the \ac{ki} itself, reconfiguration of the integral path of the \acl{LF} can be carried out solely by varying this gain, as opposed to the \ac{FIR} solution where the value of each tap must be recalculated to change the response of the system. The combination of reconfigurability and simpler hardware, make the \ac{IIR} the more desirable \ac{LF} design.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../simple_pi.pdf}
	\caption[Basic \ac{IIR} \ac{PI} loop filter]{Basic \ac{IIR} \ac{PI} loop filter.}
	\label{fig:iir_pi}
\end{figure}

>>>>

\section{Chapter Overview}
The first step in creating \iac{FPGA} based network of \acfp{ADPLL} is the design of the \ac{ADPLL} itself, which will be addressed in this chapter. The nature of \iac{FPGA} necessitates compromises in the design of a certain blocks, limiting direct transfer from \ac{ASIC} designs. In this chapter the potential designs for each individual block, or module, investigated will be explained and the case for their selection in \iac{FPGA} based \ac{ADPLL} made. A number of blocks have implemented by purely digital circuitry and as such can be transferred in their entirety from \iac{FPGA} and vice versa. However, those that will be used to emulate mixed-signal circuitry, such as the \ac{DCO} and \ac{PFD}, will be examined in greater detail. \Iac{RTL} diagram of the \ac{ADPLL} topology that will be used in the following sections is shown in Figure \ref{fig:network_adpll}.
\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{../network_adpll}
	\caption{\acs{ADPLL} designed for use in a Cartesian grid network.}
	\label{fig:network_adpll}
\end{figure}

@@

\subsection{\acs{FIR} \acl{LF}}
The \acl{FIR} implementation of the \acl{LF} will take the form presented in Figure \ref{fig:fir_pi} (a), and consists series of time delays, or ``taps'', the output of which is multiplied by a coefficient. The result of each multiplication is added together to form the filter output. Various tools, Matlab for example, can be used to compute the coefficients and number of taps required to obtain a specific filter response. Using \iac{FIR} filter is advantageous as it is always stable due to the lack of a feedback path, however a large number of taps are required to obtain a frequency response akin to that of the \ac{IIR} implementation, thus requiring an equal number of multipliers and $taps-1$ adders. Compared to \iac{IIR}, which requires just two adders, two multipliers and a flip-flop this, has a much greater hardware footprint. It is possible to condense the \ac{FIR} multiplication and addition to a single multiplier and adder, depicted in \ref{fig:fir_pi} (b), however this will take as many clock cycles to complete as there are taps. This is not viable in an \ac{ADPLL} as the filter must produce its output in a single clock cycle. In the case of any coefficient symmetry, addition can be performed prior to multiplication, thus reducing the number of multipliers for each identical coefficient.
\begin{figure}[h]
	\centering
	\subfloat[Typical FIR Filter]{{\includegraphics[width=0.6\textwidth]{../FIR_1} }}\\
	\subfloat[Single Multiplier FIR Filter]{{\includegraphics[width=0.6\textwidth]{../FIR_2} }}%
	\caption[\acs{FIR} \acs{PI} loop filter RTL diagrams]{\acs{FIR} \acs{PI} loop filter RTL diagrams.}
	\label{fig:fir_pi}
\end{figure}

@@

As the scaling factor used in this multiplier is the \ac{ki} itself, reconfiguration of the integral path of the \acl{LF} can be carried out solely by varying this gain, as opposed to the \ac{FIR} solution where the value of each tap must be recalculated to change the response of the system. The combination of reconfigurability and simpler hardware, make the \ac{IIR} the more desirable \ac{LF} design.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../simple_pi.pdf}
	\caption[Basic \acs{IIR} \acs{PI} loop filter]{Basic \acs{IIR} \acs{PI} loop filter.}
	\label{fig:iir_pi}
\end{figure}
